
---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true

---

```{r Setup}
rm(list=ls())
options(stringsAsFactors = FALSE)
options(digits=2)

Sys.setlocale("LC_TIME", "English")

if (!is.element("ggplot2",installed.packages()) ) {install.packages("ggplot2")}
library(ggplot2)

```

## Loading and preprocessing the data

```{r Loading Data}

if (!file.exists("activity.csv")) {unzip("repdata_data_activity.zip")}
dataAct <- read.csv("activity.csv")

```

## What is mean total number of steps taken per day?

```{r  Mean per Day}

##      The total number of steps taken per day

dailyData <-data.frame(date=character(length(levels(as.factor( dataAct$date)))))
dailyData$date<-unique(dataAct$date)
dailyData$tot<-tapply(dataAct$steps,dataAct$date,sum,na.rm=TRUE)

##      Histogram of the total number of steps taken each day

#png(file="stepsPerDay.png", width = 480, height = 480)
q <- qplot(dailyData$tot, xlab="Total Number of Steps")
print(q)

#dev.off()

##      The mean and median of the total number of steps taken per day     

dailyDataMean<-mean(dailyData$tot)

dailyDataMedian<-median(dailyData$tot)
```
The mean of the number of steps taken per day is 
`r dailyDataMean`  
The median of the number of steps taken per day is 
`r dailyDataMedian`  

## What is the average daily activity pattern?

```{r Average Daily}
##    plot of the 5-minute interval (x-axis) 
##    and the average number of steps taken, averaged across all days (y-axis)

avInterval<-tapply(dataAct$steps,as.factor(dataAct$interval),mean,na.rm=TRUE)
tmp<-cbind(levels(factor(dataAct$interval)),avInterval)

#png(file="dailyAverageSteps.png", width = 480, height = 480)
p<-plot (tmp[,1],tmp[,2],type="l",
         main="Daily Average No. of Steps per Time Interval",
         ylab="Av. no. of Steps",xlab="Time Interval")
#print (p)
#dev.off()
maxNoSteps<-max(as.numeric(tmp[,2]))
intervalofMax<-tmp[match(maxNoSteps,tmp[,2]),1]

```
The 5-minute interval, on average across all the days in the dataset, containning the maximum number of steps is `r intervalofMax`  

## Imputing missing values 

```{r Calculate no. of  missing values}
     
naNum <- sum(!complete.cases(dataAct))

```
total number of missing cases is `r naNum`  

Filling the missing values in the dataset is performed based on the average for that day, unless all the values of the day are NA's, than it is replaced by zero.

```{r Filling of  missing values}
##      replicate the data into a clean data frame
cleanDataAct<-dataAct

##      Calculate daily average per interval
dailyData$av <- tapply(dataAct$steps,as.factor(dataAct$date),mean,na.rm=TRUE)
##      if the average column contains NA's it is because the input to 
##      the mean function is all NA's.
##      In that case it should be 0 as the total function
dailyData$av[is.na(dailyData$av)]<-0

##      Calculate the indices of NA values in steps vector
naInd<-which(is.na(cleanDataAct$steps))
cleanDataAct$steps[naInd]<-dailyData$av[cleanDataAct$date[naInd]]

##     Plot a histogram of the total number of steps taken each day
dailyData$totClean<-tapply(cleanDataAct$steps,as.factor(cleanDataAct$date),sum)
#png(file="totalNoOfSteps.png", width = 480, height = 480)
r<- qplot (dailyData$totClean, xlab="Total Number of Daily Steps",
           binwidth = 2200, fill=I("red"))
print(r)
#dev.off()

##     The mean and median total number of steps taken per day

dailyDataMeanClean<-mean(dailyData$totClean)
dailyDataMedianClean<-median(dailyData$totClean)

```

The mean of the clean data is `r dailyDataMeanClean`  
The median of the clean data is `r dailyDataMedianClean`     

```{r difference of values without NA}

##     Do these values differ from the estimates from the first part of the 
##     assignment? 
##     ______________________________________________________________________

dailyDataMeanDIff<- dailyDataMeanClean - dailyDataMean
dailyDataMedianDiff<- dailyDataMedianClean - dailyDataMedian


```
The difference of the mean without NA's is `r dailyDataMeanDIff`   
The difference of the median without NA's is `r dailyDataMedianDiff`


```{r difference of total}
##     What is the impact of imputing missing data on the 
##     estimates of the total daily number of steps? 
dailyDataTotDIff<- sum(dailyData$totClean) - sum(dailyData$tot)

```
The impact of inputing missing data on the total daily estimates, by the method of using daily average when exists and 0 when missing for that day,  is 
`r dailyDataTotDIff`




## Are there differences in activity patterns between weekdays and weekends?

```{r difference wkdays and wkends}
##      Create a new factor variable in the dataset with two levels
##       – “weekday” and “weekend” indicating whether a given date is a weekday
##        or weekend day.
wkedays<-c("Saturday","Sunday")
cleanDataAct$dayType<-"weekday"
wkeDaysInd<-
       weekdays(as.Date(cleanDataAct$date)) %in% wkedays
cleanDataAct$dayType[wkeDaysInd]<-"weekend"
cleanDataAct$dayType <- as.factor(cleanDataAct$dayType)
l<-levels(cleanDataAct$dayType)
```
New factor variable in the dataset with two levels: `r l`

```{r difference wkdays and wkends 2}
## Make a panel plot containing a time series plot (i.e. type = "l") 
## of the 5-minute interval (x-axis) and the average number of steps taken, 
## averaged across all weekday days or weekend days (y-axis).


#prepare dataframe cleanAV for plotting 
#
## first cleanAv has the average number of steps taken, 
## averaged across all weekday days 
## and cleanAve has the average number of steps taken, 
## averaged across  all weekend days.
## Than they are combined in cleanAv
#
cleanAv<-data.frame(interval=integer
                    (length(levels(as.factor( cleanDataAct$interval)))))
cleanAv$interval<-levels(factor(cleanDataAct$interval))
cleanAv$wType<-as.factor("weekday")
cleanAve<-data.frame(interval=integer
                     (length(levels(as.factor( cleanDataAct$interval)))))
cleanAve$interval<-levels(factor(cleanDataAct$interval))
cleanAve$wType<-as.factor("weekend")

#prepare weekday data for ploting in cleanAv dataframe
df<-data.frame(wSteps=
                       numeric(length(which(cleanDataAct$dayType == "weekday"))))
df$wSteps<-cleanDataAct$steps[which(cleanDataAct$dayType=="weekday")]
df$wInterval<-cleanDataAct$interval[which(cleanDataAct$dayType=="weekday")]

cleanAv$av<-tapply(df$wSteps,as.factor(df$wInterval),mean,na.rm=TRUE)

#prepare weekend data for ploting in auxiliary
de<-data.frame(wSteps=
                       numeric(length(which(cleanDataAct$dayType == "weekend"))))
de$wSteps<-cleanDataAct$steps[which(cleanDataAct$dayType=="weekend")]
de$wInterval<-cleanDataAct$interval[which(cleanDataAct$dayType=="weekend")]

cleanAve$av<-tapply(de$wSteps,as.factor(de$wInterval),mean,na.rm=TRUE)

cleanAv<-rbind(cleanAv, cleanAve)


# plot
#png(file="difWeekdaysWeekends.png", width = 480, height = 480)
#par(mar = c(4, 4, 4, 4)) 

f<-ggplot(cleanAv,aes(interval, av ))
f<-f+geom_line(aes(group=wType, color = wType))
        
f<-f+facet_grid(wType~.)
f<-f+labs(y="Number of Steps")
      
print (f)
#dev.off()
```

